A Formal Model of Checked C

### 2023-01-09 sandp23/formal.tex
Adding

Multi-word values represent arrays and null-terminated arrays, which include both the type of elements ($\tau$) and a bound ($\bvar$).
The bounds can be either integer literals $n$, expressions $x + n$, or an explicit specification of lower and upper bound,~\ie $(b_l,b_h)$.
We denote null-terminated arrays using $\kappa$, which is $nt$ for null-terminated arrays, and $\cdot$ otherwise (we elide $\cdot$ when writing types).
The well-formed requirements of these types are presented in~\fig{fig:wftypesandbounds} (Appendix).
An example representation of an array and null-terminated array in~\lang syntax is shown below:
\[\hspace*{-0.5em}
\begin{array}{l}
\begin{array}{rcl}
$\code{_t_Array_ptr<}$\tau$\code{> : count(}$n$\code{)}$
&\Leftrightarrow& \tarrayptr{0}{n}{\tau}{\tmode}
\\[0.2em]
$\code{_NT_Array_ptr<}$\tau$\code{> : count(}$n$\code{)}$
&\Leftrightarrow& \tntarrayptr{0}{n}{\tau}{\cmode}
\end{array}
\end{array}
\]
To simplify the representation, we write $\tptr{\tarrayb{b}{\tau}}{\cmode}$ to mean
$\tptr{\tarray{0}{b}{\tau}}{\cmode}$, so the above examples could
be rewritten $\tptr{\tarrayb{n}{\tau}}{\cmode}$ and
$\tptr{\tntarrayb{n}{\tau}}{\cmode}$, respectively.

Function types are represented using a dependent function declarations,~\ie $\tfun{\overline{x}}{\overline{\tau}}{\tau}$,
where $\overline{x}$ represents a list of \tint{} type variables that bind bound variables appearing in $\overline{\tau}$ and $\tau$.
An example of a function pointer type is shown below:
\[\hspace*{-0.5em}
\begin{array}{l}
$\code{_t_ptr<(int)(_t_NT_Array_ptr<}$\tau$\code{> : count(}$n$\code{),}$\\
\qquad\qquad$\code{_NT_Array_ptr<}$\tau$\code{>: count(}$n$\code{))>}$
\\[0.2em]
\Leftrightarrow\;\; $\tptr{(\tfun{n}{ \tntarrayptr{0}{n}{\tau}{\tmode} \times \tntarrayptr{0}{n}{\tau}{\tmode}}{\tint})}{\tmode}$
\end{array}
\]
Note that the well-formed requirement (\fig{fig:wffunctions} in Appendix) of a function type ensures that all variables in $\overline{\tau}$ and $\tau$ are bounded by $\overline{x}$.
\aravind{The syntax of well-formed requirements for functions is incorrect. It looks like the forall qualifier is missing.}

@@

\aravind{Why we have only function pointers in function call?}

>>>>

Multi-word values represent arrays and null-terminated arrays, which include both the type of elements ($\tau$) and a bound ($\bvar$).
The bounds can be either integer literals $n$, expressions $x + n$, or an explicit specification of lower and upper bound,~\ie $(b_l,b_h)$.
We denote null-terminated arrays using $\kappa$, which is $nt$ for null-terminated arrays, and $\cdot$ otherwise (we elide $\cdot$ when writing types).
An example representation of an array and null-terminated array in~\lang syntax is shown below:
\[\hspace*{-0.5em}
\begin{array}{l}
\begin{array}{rcl}
$\code{_t_Array_ptr<}$\tau$\code{> : count(}$n$\code{)}$
&\Leftrightarrow& \tarrayptr{0}{n}{\tau}{\tmode}
\\[0.2em]
$\code{_NT_Array_ptr<}$\tau$\code{> : count(}$n$\code{)}$
&\Leftrightarrow& \tntarrayptr{0}{n}{\tau}{\cmode}
\end{array}
\end{array}
\]
To simplify the representation, we write $\tptr{\tarrayb{b}{\tau}}{\cmode}$ to mean
$\tptr{\tarray{0}{b}{\tau}}{\cmode}$, so the above examples could
be rewritten $\tptr{\tarrayb{n}{\tau}}{\cmode}$ and
$\tptr{\tntarrayb{n}{\tau}}{\cmode}$, respectively.

\emph{Disallowing unsafe types.}
The well-formed requirements of these types are presented in~\fig{fig:wftypesandbounds} (Appendix).
These requirements add additional restrictions for type creation and restrict the creation of unsafe types.
For instance, we cannot create a tainted array whose elements are checked pointers,~\eg Although syntactically valid, the type~\code{_t_Array_ptr<_Ptr<int>>} is~\emph{not} well-formed and is not allowed in~\lang.
However, we can have a checked array whose elements are tainted pointers,~\eg~\code{_t_Array_ptr<_t_Ptr<int>>} is a valid type.

Function types are represented using a dependent function declarations,~\ie $\tfun{\overline{x}}{\overline{\tau}}{\tau}$,
where $\overline{x}$ represents a list of \tint{} type variables that bind bound variables appearing in $\overline{\tau}$ and $\tau$.
An example of a function pointer type is shown below:
\[\hspace*{-0.5em}
\begin{array}{l}
$\code{_t_ptr<(int)(_t_NT_Array_ptr<}$\tau$\code{> : count(}$n$\code{),}$\\
\qquad\qquad$\code{_NT_Array_ptr<}$\tau$\code{>: count(}$n$\code{))>}$
\\[0.2em]
\Leftrightarrow\;\; $\tptr{(\tfun{n}{ \tntarrayptr{0}{n}{\tau}{\tmode} \times \tntarrayptr{0}{n}{\tau}{\tmode}}{\tint})}{\tmode}$
\end{array}
\]
The function type also has well-formed requirements (\fig{fig:wffunctions} in Appendix), which disallows nesting checked or unchecked pointers inside tainted pointers.
Furthermore, these requirements also ensure that all variables in $\overline{\tau}$ and $\tau$ are bounded by $\overline{x}$.
\aravind{The syntax of well-formed requirements for functions is incorrect. It looks like the forall qualifier is missing.}

@@

\aravind{Why we have only function pointers in function call?}
\aravind{Talk about the ret expression after adding it in the syntax.}
